ICM testing

### 2022-09-02 Scripts/ASR_stocks.Rmd
ICES stocks completed
I have run the analysis on the ICES stocks with 250 replicates per run, took about 6 hours to run for 46 stocks.

The stocks chosen for testing were those with the most complete data.

Some catch information was missing for certain years. These years are marked with "?" in the plots that follow. The median annual catch was assumed for missing years.

Catch data for _Sprattus sprattus_ (European sprat) was manually converted to be on the 1000s scale, in order to align with abundance data.

@@

source("./functions/backwards_sim.R")

# Choose 5 ICES stocks that we have the necessary data for
ASR <- read_xlsx("../Data/ASR_2018.xlsx", sheet = "ICES")
datatypes <- unique(gsub(x = names(ASR), pattern = "[^a-zA-Z]", replacement=""))
# we want:
# Year, Num, WA, Catch, AM, NM, StockID, Management, Area, Order, Family, Genus, Species
ASRdat <- ASR[,c(grep(x=names(ASR), "Num"),
                 grep(x=names(ASR), "WA"),
                 grep(x=names(ASR), "Catch"),
                 grep(x=names(ASR), "AM"),
                 grep(x=names(ASR), "NM"))]
ASRdat <- apply(X = ASRdat, 2, as.numeric)
ASRsp <- ASR[, which(!1:length(names(ASR)) %in% grep(x=names(ASR), ".", fixed=T))]
ASR_trim <- cbind(ASRsp, ASRdat)

@@

ASR_long <- ASR_trim %>%
  pivot_longer(!c("Management", "Area", "Order", "Family", "Genus", "Species", "Stock", "Year")) %>%
  separate(col=name, into=c("type", "age"), sep = "\\.")

# summary plots for ALL stocks
for (i in unique(ASR_long$Stock)) {
  ASR_sub <- ASR_long %>%
    select(Year, Stock, type, age, value) %>%
    filter(Stock==i)
  maxage <- max(as.numeric(ASR_sub$age[!is.na(ASR_sub$value)]))
  maxYear <- max(ASR_sub$Year[!is.na(ASR_sub$value)])
  minYear <- min(ASR_sub$Year[!is.na(ASR_sub$value)])
  forplot <- ASR_sub[as.numeric(ASR_sub$age) < (maxage+1) & ASR_sub$Year %in% minYear:maxYear,]
  p1 <- ggplot() + geom_line(data=forplot, aes(Year, value)) + facet_grid(type~as.numeric(age), scales="free_y") +
        geom_vline(data=forplot[is.na(forplot$value),], aes(xintercept=Year), colour="red") + facet_grid(type~as.numeric(age), scales="free_y") +
        theme_bw() + 
        ggtitle(i) +
    guides(x =  guide_axis(angle = 90)) 
  print(p1)

  if(!dir.exists(paste0("../Figures/", i))) dir.create(paste0("../Figures/", i))

  png(filename=paste0("../Figures/", i, "/data_summary_", i, ".png"), height=4, width=12, res=400, units="in")
  print(p1)
  dev.off()
}

Stocks <- ASR_long %>%
  filter(!is.na(value)) %>%
  group_by(Stock, type) %>%
  summarize(count=length(unique(value))) %>%
  group_by(Stock) %>%
  summarize(types=length(unique(type))) %>%
  filter(types==5) %>%
  select(Stock)

ASR_stocks <- ASR_long %>%
  filter(Stock %in% Stocks$Stock) %>%
  filter(!is.na(value)) %>%
  arrange(Stock, Year, type, as.numeric(age))

Stocks <- ASR_stocks %>%
  group_by(Stock, Species, type) %>%
  summarize(ages=length(unique(age)),
            years=length(unique(Year))) %>%
  arrange(-years, -ages) %>%
  filter(!Species=="morhua") %>%
  distinct(Stock) %>%
  pull(Stock)

print(Stocks)

# Bring in the data
# Handy import function from the rio package... tricky part was getting the right filename here...
for(i in Stocks){
  print(i)
  ASR_sub <- ASR_long %>%
    select(Year, Stock, type, age, value) %>%
    filter(Stock==i)

@@

  age.mat <- ASR_sub %>% filter(type=="AM") %>% rename(AM=value) %>% select(-type)
  nat.mort <-  ASR_sub %>% filter(type=="NM") %>% rename(NM=value) %>% select(-type)
  abund <- ASR_sub %>% filter(type=="Num") %>% rename(Num=value) %>% select(-type)
  weight.age <- ASR_sub %>% filter(type=="WA") %>% rename(WA=value) %>% select(-type)
  removals <- ASR_sub %>% filter(type=="Catch") %>% rename(Catch=value) %>% select(-type)

  if(i=="ICES-WGBFAS_BS 22-32_Sprattus_sprattus") removals$Catch <- removals$Catch/1000

@@

  rem <- data %>% group_by(Year) %>% summarize(rem=sum(Catch)) %>% pull(rem)

@@

  #rowSums(removals[,-which(names(removals) %in% c("Year", "Stock", "type"))], na.rm=T)
  years <- data %>% pull(Year) %>% unique() %>% sort()
  N.end <- sum(data[data$Year==max(years),]$Num)
  #N.end <- rowSums(abund[nrow(abund),-which(names(abund) %in% c("Year", "Stock", "type"))], na.rm=T)
  #vpa.abund <- rowSums(abund[,-which(names(abund) %in% c("Year", "Stock", "type"))], na.rm=T)
  vpa.abund <- data %>% group_by(Year) %>% summarize(vpa=sum(Num)) %>% pull(vpa)

  # The real mx matrix, recruits produced per individual in each age class... Not perfect as I need to offset recruits/ssb, but close enough for the moment..
  #minage
  minage <- min(as.numeric(data$age))
  #maxage
  maxage <- max(as.numeric(data$age))
  #recruits
  annual <- data.frame(Year=data$Year[data$age==minage], recruits=data$Num[data$age==minage])
  #ssn
  data$ssn <- data$Num * data$AM
  #ssb
  data$ssb <- data$ssn * data$WA
  #tot.ssb
  annual <- data %>% group_by(Year) %>% summarize(tot.ssb = sum(ssb)) %>% left_join(annual)
  #r.p.ssb
  annual$r.p.ssb <- annual$recruits/annual$tot.ssb
  # recs.per.age
  data <- left_join(data, annual)
  data$recs.per.age <- data$ssb*data$r.p.ssb
  # mx
  data$mx <- data$recs.per.age/data$ssn/2 # Moms only! Dividing by around 8 really nails it for NS cod for whatever reason :-)
  # Something is wrong with the decline rate method, but the exponential and logistic are working pretty... pretty... pretty good...

  age.mat <- data %>% select("Year", "age", "AM") %>% pivot_wider(names_from=age, values_from = AM) %>% select(-Year)
  prop.nat.mort <- data %>% select("Year", "age", "prop.nat.mort") %>% pivot_wider(names_from=age, values_from = prop.nat.mort) %>% select(-Year)
  weight.age <- data %>% select("Year", "age", "WA") %>% pivot_wider(names_from=age, values_from = WA) %>% select(-Year)
  mx <- data %>% select("Year", "age", "mx") %>% pivot_wider(names_from=age, values_from = mx) %>% select(-Year) %>% as.data.frame()

  tst <- icm.sim(years = years,
                 mat.age = age.mat,
                 nm = prop.nat.mort,
                 w.age = weight.age,
                 ages = minage:maxage,
                 rems = rem,
                 fecund = mx,
                 N.end = N.end,
                 pop.model = 'exponential',
                 n.sims = 5,
                 sd.mat = 0.5,
                 sd.nm = 0.5,
                 sd.wt = 0.5,
                 sd.fecund = 0.5)

@@

  png(filename=paste0("../Figures/", i, "/runs_vs_vpa_", i, ".png"), height=4, width=6, res=400, units="in")
  print(p2)
  dev.off()
  png(filename=paste0("../Figures/", i, "/runs_vs_vpa_ribbon_", i, ".png"), height=4, width=6, res=400, units="in")
  print(p3)
  dev.off()
  png(filename=paste0("../Figures/", i, "/growth_", i, ".png"), height=4, width=6, res=400, units="in")
  print(p4)
  dev.off()
}
```

>>>>

The data is now avilable to test this on at least 81 stocks.  All of the NE Atlantic stocks (ICES) have been through the QA/QC process and we are happy with those data (Sept 1, 2022).  Let's see if we can get this to work on all the stocks with sufficient data...

@@

source("D:/Github/ICM/Scripts/functions/backwards_sim.R")

# Choose 5 ICES stocks that we have the necessary data for
#ASR <- read_xlsx("../Data/ASR_2018.xlsx", sheet = "ICES")
ASR <- read.csv("D:/Github/ICM/Data/ICM_data_NE_atlantic_stocks.csv")
datatypes <- unique(gsub(x = names(ASR), pattern = "[^a-zA-Z]", replacement=""))

# Replace 0's in Num.tot with NA so the rest of this works as Freya designed it to...
ASR$Num.tot[ASR$Num.tot == 0] <- NA
# we want:
# Year, Num, WA, Catch, AM, NM, StockID, Management, Area, Order, Family, Genus, Species
ASRdat <- ASR[,c(grep(x=names(ASR), "Num"),
                 grep(x=names(ASR), "WA"),
                 grep(x=names(ASR), "Catch"),
                 grep(x=names(ASR), "AM"),
                 grep(x=names(ASR), "NM"))]
# Making all the data numberic that should be numeric
ASRdat <- apply(X = ASRdat, 2, as.numeric)
# Getting the species info back
ASRsp <- ASR[, which(!1:length(names(ASR)) %in% grep(x=names(ASR), ".", fixed=T))]
# And binding it all back together
ASR_trim <- cbind(ASRsp, ASRdat)

@@

ASR_long <- ASR_trim %>%
  pivot_longer(!c("Management", "Area", "Order", "Family", "Genus", "Species", "Stock", "Year","Model","Case","Notes")) %>%
  separate(col=name, into=c("type", "age"), sep = "\\.")

# summary plots for ALL stocks
for (i in unique(ASR_long$Stock)) {
  ASR_sub <- ASR_long %>%
    dplyr::select(Year, Stock, type, age, value) %>%
    dplyr::filter(Stock==i)
  maxage <- max(as.numeric(ASR_sub$age[!is.na(ASR_sub$value)]))
  maxYear <- max(ASR_sub$Year[!is.na(ASR_sub$value)])
  minYear <- min(ASR_sub$Year[!is.na(ASR_sub$value)])
  forplot <- ASR_sub[as.numeric(ASR_sub$age) < (maxage+1) & ASR_sub$Year %in% minYear:maxYear,]
  p1 <- ggplot() + geom_line(data=forplot, aes(Year, value)) + facet_grid(type~as.numeric(age), scales="free_y") +
        geom_vline(data=forplot[is.na(forplot$value),], aes(xintercept=Year), colour="red") + facet_grid(type~as.numeric(age), scales="free_y") +
        theme_bw() + 
        ggtitle(i) +
    guides(x =  guide_axis(angle = 90)) 
  print(p1)

  if(!dir.exists(paste0("D:/Github/ICM/Figures/", i))) dir.create(paste0("D:/Github/ICM/Figures/", i))

  png(filename=paste0("D:/Github/ICM/Figures/", i, "/data_summary_", i, ".png"), height=4, width=12, res=400, units="in")
  print(p1)
  dev.off()
}

Stocks <- ASR_long %>%
                    dplyr::filter(!is.na(value)) %>%
                    dplyr::group_by(Stock, type) %>%
                    dplyr::summarize(count=length(unique(value))) %>%
                    dplyr::group_by(Stock) %>%
                    dplyr::summarize(types=length(unique(type))) %>%
                    dplyr::filter(types==5) %>%
                    dplyr::select(Stock)

ASR_stocks <- ASR_long %>%
                        dplyr::filter(Stock %in% Stocks$Stock) %>%
                        dplyr::filter(!is.na(value)) %>%
                        dplyr::arrange(Stock, Year, type, as.numeric(age))

# Stocks <- ASR_stocks %>%
#                       dplyr::group_by(Stock, Species, type) %>%
#                       dplyr::summarize(ages=length(unique(age)),
#                                 years=length(unique(Year))) %>%
#                       dplyr::arrange(-years, -ages) %>%
#                       dplyr::filter(!Species=="morhua") %>%
#                       dplyr::distinct(Stock) %>%
#                       dplyr::pull(Stock)
# 
# print(Stocks)
Stocks <- Stocks$Stock
# Going to remove Sebastes norvegicus because we can't get fecundity as the only data we have is total numbers.
Stocks <- Stocks[Stocks != "ICES-AFWG_DEEP1-2_Sebastes_norvegicus"]
# Esmarki doesn't work great as it is assessed using a quarterly modle.
Stocks <- Stocks[Stocks != "ICES-WGNSSK_NS 4-3aN_Trisopterus_esmarkii"]
# Bring in the data
# Handy import function from the rio package... tricky part was getting the right filename here...
for(i in Stocks){
  print(i)
  ASR_sub <- ASR_long %>%
    dplyr::select(Year, Stock, type, age, value) %>%
    dplyr::filter(Stock==i)

@@

  age.mat <- ASR_sub %>% dplyr::filter(type=="AM") %>% dplyr::rename(AM=value) %>% dplyr::select(-type)
  nat.mort <-  ASR_sub %>% dplyr::filter(type=="NM") %>% dplyr::rename(NM=value) %>% dplyr::select(-type)
  abund <- ASR_sub %>% dplyr::filter(type=="Num") %>% dplyr::rename(Num=value) %>% dplyr::select(-type)
  weight.age <- ASR_sub %>% dplyr::filter(type=="WA") %>% dplyr::rename(WA=value) %>% dplyr::select(-type)
  removals <- ASR_sub %>% dplyr::filter(type=="Catch") %>% dplyr::rename(Catch=value) %>% dplyr::select(-type)

 # if(i=="ICES-WGBFAS_BS 22-32_Sprattus_sprattus") removals$Catch <- removals$Catch/1000

@@

  rem <- data %>% dplyr::group_by(Year) %>% dplyr::summarize(rem=sum(Catch,na.rm=T)) %>% dplyr::pull(rem)

@@

  #rowSums(removals[,-which(names(removals) %in% c("Year", "Stock", "type"))], na.rm=T)
  years <- data %>% pull(Year) %>% unique() %>% sort()
  N.end <- sum(data[data$Year==max(years),]$Num)
  #N.end <- rowSums(abund[nrow(abund),-which(names(abund) %in% c("Year", "Stock", "type"))], na.rm=T)
  #vpa.abund <- rowSums(abund[,-which(names(abund) %in% c("Year", "Stock", "type"))], na.rm=T)
  vpa.abund <- data %>% dplyr::group_by(Year) %>% dplyr::summarize(vpa=sum(Num)) %>% pull(vpa)

  # The real mx matrix, recruits produced per individual in each age class... Not perfect as I need to offset recruits/ssb, but close enough for the moment..
  #minage
  minage <- min(as.numeric(data$age))
  #maxage
  maxage <- max(as.numeric(data$age))
  #recruits
  annual <- data.frame(Year=data$Year[data$age==minage], recruits=data$Num[data$age==minage])
  #ssn
  data$ssn <- data$Num * data$AM
  #ssb
  data$ssb <- data$ssn * data$WA
  #tot.ssb
  annual <- data %>% group_by(Year) %>% summarize(tot.ssb = sum(ssb)) %>% left_join(annual)
  #r.p.ssb
  annual$r.p.ssb <- annual$recruits/annual$tot.ssb
  # recs.per.age
  data <- left_join(data, annual)
  data$recs.per.age <- data$ssb*data$r.p.ssb
  # mx
  data$mx <- data$recs.per.age/data$ssn/2 # Moms only! Dividing by around 8 really nails it for NS cod for whatever reason :-)
  data$mx[is.nan(data$mx)] <- 0 # if we don't have any spawners in an age class in a year their fecundity is 0
  # Something is wrong with the decline rate method, but the exponential and logistic are working pretty... pretty... pretty good...

  age.mat <- data %>% dplyr::select("Year", "age", "AM") %>% pivot_wider(names_from=age, values_from = AM) %>% dplyr::select(-Year)
  prop.nat.mort <- data %>% dplyr::select("Year", "age", "prop.nat.mort") %>% pivot_wider(names_from=age, values_from = prop.nat.mort) %>% dplyr::select(-Year)
  weight.age <- data %>% dplyr::select("Year", "age", "WA") %>% pivot_wider(names_from=age, values_from = WA) %>% dplyr::select(-Year)
  mx <- data %>% dplyr::select("Year", "age", "mx") %>% pivot_wider(names_from=age, values_from = mx) %>% dplyr::select(-Year) %>% as.data.frame()

  tst <- icm.sim(years = years,
                 mat.age = age.mat,
                 nm = prop.nat.mort,
                 w.age = weight.age,
                 ages = minage:maxage,
                 rems = rem,
                 fecund = mx,
                 N.end = N.end,
                 pop.model = 'exponential',
                 n.sims = 1000,
                 sd.mat = 0.5,
                 sd.nm = 0.5,
                 sd.wt = 0.5,
                 sd.fecund = 0.5)

@@

  png(filename=paste0("D:/Github/ICM/Figures/", i, "/runs_vs_vpa_", i, ".png"), height=4, width=6, res=400, units="in")
  print(p2)
  dev.off()
  png(filename=paste0("D:/Github/ICM/Figures/", i, "/runs_vs_vpa_ribbon_", i, ".png"), height=4, width=6, res=400, units="in")
  print(p3)
  dev.off()
  png(filename=paste0("D:/Github/ICM/Figures/", i, "/growth_", i, ".png"), height=4, width=6, res=400, units="in")
  print(p4)
  dev.off()
}
```
